A new machine-learning-based prediction of survival in patients with end-stage liver disease

### 2022-11-30 analysis/article.Rmd
feat: add MELD 3.0 and adapt to reviewer's suggestions

---
# R markdown template taken from
# Dan Holmes <dtholmes@mail.ubc.ca>
# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
title: "A new machine-learning-based prediction of survival in patients with end-stage liver disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Objectives**:
      The shortage of grafts for liver transplantation requires
      risk stratification and adequate allocation rules.
      This study aims to improve the model of end-stage liver disease (MELD)
      score for 90-day mortality prediction with the help of different
      machine-learning algorithms.

@@

  **Results**:
      Penalized regression algorithms yielded the highest prediction
      performance in our machine-learning algorithm benchmark.
      In favor of a simpler model we chose the least absolute shrinkage and
      selection operator (lasso) regression.
      Beside the classical MELD variables international normalized ratio (INR)
      and bilirubin, the lasso regression selected cystatin C over creatinine,
      as well IL-6, total protein, and cholinesterase.
      The new model offers improved discrimination and calibration over MELD
      and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.

  **Conclusion**:
      We provide a new machine-learning-based model of end-stage liver disease
      that outperforms the classical MELD score for 90-day survival prediction.

@@

tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_RCV)
```

@@

Liver cirrhosis is the terminal result of the fibrotic remodeling
of liver tissue due to chronic damage.
This end-stage of liver disease is usually irreversible, and
the only available therapy is liver transplantation.
However, the shortage of grafts for transplantation from deceased donors
requires risk stratification and precise and fair allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine, and
the international normalized ratio (INR).
The MELD score has been extended by the sodium level
(MELD-Na score) because this was found to be an important additional
risk factor in liver cirrhosis [@kim2008; @optn_policies2021].
The MELD was initially developed to predict the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
It was subsequently revalidated to predict mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score,
creatinine and INR are highly dependent on the laboratory methods used
[@trotter2004; @cholongitas2007].
In addition women are disadvantaged [@sealock2022].
Patients with identical disease states can have very different MELD scores and
thus receive different priority levels on the liver transplantation waiting
list.
Furthermore, for acute-on-chronic liver failure, for example, the MELD score
often underestimates the mortality risk [@hernaez2020].

@@

As shown by the MELD-Plus risk scores, better predictive scores often need more
variables and are more complicated. To reduce the risk of overlooking or
incorrectly calculating and interpreting the results,
clinical decision support systems may be used and could improve patient safety.
The research project on digital laboratory medicine (AMPEL) develops a
clinical decision support system based on laboratory diagnostics that should
support clinical practitioners in interpreting laboratory results and taking
the necessary medical interventions [@eckelt2020].

This study aims to find clinical and laboratory values that improve on the risk
stratification for liver transplantation of the classical MELD, MELD-Na, and
MELD-Plus scores and that might be implemented as part of the
AMPEL clinical decision support system.

@@

The MELD-Plus7 risk score was calculated as described in @kartoun2017:
$$
\begin{aligned}
L = 8.53499496 &+ \\
    2.59679650 &* \log_{10}(1 + creatinine [mg/dL]) + \\
    2.06503238 &* \log_{10}(1 + bilirubin [mg/dL]) + \\
    2.99724802 &* \log_{10}(1 + INR) - \\
    6.47834101 &* \log_{10}(1 + sodium [mmol/L]) - \\
    6.34990436 &* \log_{10}(1 + albumin [g/L]) + \\
    1.92811726 &* \log_{10}(1 + wbc [th/cumm]) + \\
    0.04070442 &* age [years]
\end{aligned}
$$

@@

## Data processing and machine learning

@@

The final model was compared against the MELD, MELD-Na, and MELD-Plus7 scores
using the area under the time-dependent receiver operating characteristic curve
(AUROC) based on the nonparametric inverse probability of censoring weighting
estimator (IPCW) [@blanche2013, @R-timeROC].
Testing was applied as described in @blanche2013.
A $p < 0.05$ was considered a statistically significant difference.

@@

Most of the patients presented with cirrhosis
(`r .tbl1("Cirrhosis")`).
The remaining patients had acute liver failure
(`r .tbl1("ALF")`) or
chronic liver insufficiency
(`r .tbl1("CLI")`).
At evaluation time, the median (IQR) MELD and MELD-Na scores were
`r .tbl1("ScoreMeldUnos")` and
`r .tbl1("ScoreMeldNaUnos")`, respectively.
Within 90 days, `r .tbl1("M90")` patients had died.
In nearly two thirds of all patients
(`r .tbl1("Ethyltoxic")`), alcohol was the main cause
of their end-stage liver disease.
The remaining patients had viral infections -
hepatitis B (`r .tbl1("HBV")`) or C (`r .tbl1("HCV")`) -
autoimmune autoimmune (`r .tbl1("AIH")`),
primary sclerosing cirrhosis (`r .tbl1("PSC")`),
primary biliary cirrhosis (`r .tbl1("PBC")`), unknown causes
(non-alcoholic steatohepatitis (`r .tbl1("NASH")`), or
cryptogenic hepatitis (`r .tbl1("Cryptogenic")`)).
The most common complications among all patients were gastrointestinal
bleedings in `r .tbl1("GIB")` patients, followed by
hepatocellular carcinoma (HCC) in `r .tbl1("HCC")` and
spontaneous bacterial peritonitis (SBP) in `r .tbl1("SBP")` patients.
Just `r .tbl1("Dialysis")` patients required renal dialysis.

@@

According to @wiesner2003, we divided our cohort into
`r nlevels(labelled_meld_data$MeldCategory)` MELD-score risk
categories, with `r enum(table(labelled_meld_data$MeldCategory))` patients,
respectively.
As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality),
our observed 90-day mortality was
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times higher than the predicted one [@vanderwerken2021].
Only in the group with the lowest MELD scores was the mortality much lower.

```{r tbl-observed-vs-expected-mortality, results = "asis", echo = FALSE, message = FALSE}
tbl_observed_vs_expected_mortality$ObservedMortality <-
    tbl_observed_vs_expected_mortality$ObservedMortality * 100
tbl_observed_vs_expected_mortality$ExpectedMortality <-
    tbl_observed_vs_expected_mortality$ExpectedMortality * 100
knitr::kable(
    cbind(
        row.names(tbl_observed_vs_expected_mortality),
        tbl_observed_vs_expected_mortality
    ),
    row.names = FALSE,
    col.names =
        c(
            "MELD category",
            "Observed deaths (n)",
            "Expected deaths (n)",
            "Standardized mortality ratio (SMR)",
            "Observed mortality (%)",
            "Expected mortality (%)"
        ),
    caption = paste0(
        "Observed vs. MELD-expected 90-day mortality. ",
        "MELD mortality values are taken from @vanderwerken2021 and ",
        "the mean value was calculated for each MELD category introduced ",
        "in @wiesner2003.",
        "All patients censored before day 90 are ignored for ",
        " the calculation of the MELD-expected deaths. ",
        "SMR, Standardized mortality ratio = observed deaths/expected deaths."
    ),
    digits = 1
)
```

@@

```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve. Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, and MELD-Plus7, as described in [@blanche2013]."}
col <- viridisLite::viridis(5)
m <- list(
    AMELD = timeROC_RCV,
    MELD = timeROC_MELD,
    "MELD-Na" = timeROC_MELDNa,
    "MELD-Plus7" = timeROC_MELDPlus7
)
invisible(roc <- plot_surv_roc(m, timepoint = 90, col = col[-2]))
roc <- roc[order(-roc)]
```

@@

AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) of all competitors.
However, after correction for multiple testing, the difference against
the second-best model, `r names(roc)[2]`
(AUROC: `r round(roc[2], 3)`), is not statistical significant
(`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
(see Table \@ref(tab:roctest) for complete comparison).

```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na, and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival. The error bars correspond to the 95% confidence interval of the survival estimate.'), echo = FALSE}
par(cex = 0.9)
plot(
    bootrcv,
    what = "calibration",
    main = "90-day survival",
    legend = FALSE
)

ps <- lapply(
    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
    function(p) {
        ctpnts <- cutpoints(p, n = ameldcfg$m)
        f <- cut(p, ctpnts, include.lowest = TRUE)
        list(
            predicted = groupmean(p, f = f),
            observed = observed_survival(
                amelddata$y, f = f, times = ameldcfg$times
            )
        )
    }
)
names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")

@@

In this retrospective analysis of `r n_analysed` consecutive patients
who were recruited during the evaluation process for liver transplantation at
University Hospital Leipzig, we developed a new model to predict
90-day survival probability. Our new model, AMELD, outperforms the MELD-Na
score and the recently published MELD-Plus7 risk score.

@@

It has been shown that women receive median lower scores in MELD-Na and
therefore are less likely to receive an urgent listing for liver transplantation
[@sealock2022].
To address sexual differences, in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017] in our study.
This transformation also yields comparable laboratory measurements
across different clinical laboratories and methods, as well as reducing
the influence of age- and sex-related differences,
which could improve the objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample-dependent scaling prior to model development.
Both should increase the external applicability of our approach.

@@

The latter is an example where results from machine-learning can aid in
finding clinically meaningful patterns and guide further research.
Systemic inflammatory response syndrome is an
independent risk factor for poor outcomes in patients with cirrhosis
[@thabut2007].
However, neither MELD nor MELD-Na took inflammation into account.
By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
excluded, the algorithms select CRP as one of the most important variables,
which highlights the important role of inflammation in the prognosis of
patients with end-stage liver disease.
We and other researchers have already shown that adding inflammatory markers,
namely IL-6 or CRP, to the classical MELD score increases the accuracy of the
90-day mortality prediction [@remmler2018; @cervoni2016].

@@

With a median MELD score of `r .tbl1("ScoreMeldUnos", pattern = "{median}")`,
our cohort was healthier than those in previous studies that reported
a score of 14-15 [@kim2008; @kartoun2017].
However, the 90-day mortality rate of `r .tbl1("M90")` was comparable to
the mortality rates seen in earlier studies ranging from 6% to 30%
[@malinchoc2000; @kim2008; @kartoun2017].
Although our cohort was healthier, with a medium mortality rate, the classical
MELD score underestimates the observed mortality by
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times.
Unfortunately, our sample size was low in the high-risk patient category,
resulting in a poor but better than MELD-based prediction performance by our
AMELD for very ill patients.

@@

In conclusion, we have developed a new model to predict the 90-day survival of
patients during the evaluation process for liver transplantation.
Our model, AMELD, extends the classical MELD predictors INR,
(direct) bilirubin, and cystatin C (instead of creatinine) by using the
synthesis parameters total protein and cholinesterase and,
most importantly, the inflammatory marker IL-6.
Using these six laboratory measurements, AMELD
outperforms the established MELD-Na and the recently published
MELD-Plus7 risk scores.
For wider adoption, a prospective multi-center study for model
recalibration and validation is needed.

>>>>

---
# R markdown template taken from
# Dan Holmes <dtholmes@mail.ubc.ca>
# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
title: "A new machine-learning-based prediction of survival in patients with end-stage liver disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD 3.0, MELD-Plus7.
abstract: >
  **Objectives**:
      The shortage of grafts for liver transplantation requires
      risk stratification and adequate allocation rules.
      This study aims to improve the model of end-stage liver disease (MELD)
      score for 90-day mortality prediction with the help of different
      machine-learning algorithms.

@@

  **Results**:
      Penalized regression algorithms yielded the highest prediction
      performance in our machine-learning algorithm benchmark.
      In favor of a simpler model we chose the least absolute shrinkage and
      selection operator (lasso) regression.
      Beside the classical MELD variables international normalized ratio (INR)
      and bilirubin, the lasso regression selected cystatin C over creatinine,
      as well IL-6, total protein, and cholinesterase.
      The new model offers improved discrimination and calibration over MELD
      and MELD with sodium (MELD-Na), MELD 3.0, or the MELD-Plus7 risk score.

  **Conclusion**:
      We provide a new machine-learning-based model of end-stage liver disease
      that incorporates synthesis and inflammatory markers and
      could improve the classical MELD score for 90-day survival prediction.

@@

tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_MELD3)
tar_load(timeROC_RCV)
```

@@

Liver cirrhosis is the terminal result of the fibrotic remodeling
of liver tissue due to chronic damage.
This end-stage of liver disease is usually irreversible, and
the only available therapy is liver transplantation.
However, the shortage of grafts for transplantation from deceased donors
requires risk stratification and precise and fair allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine, and
the international normalized ratio (INR).
The MELD score has been extended by the sodium level
(MELD-Na score) because this was found to be an important additional
risk factor in liver cirrhosis [@kim2008; @optn_policies2021].
Recently MELD 3.0 was introduced adding a factor for female sex, albumin and
interactions between bilirubin and sodium and between albumin and creatinine.
The MELD was initially developed to predict the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
It was subsequently revalidated to predict mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score,
creatinine and INR are highly dependent on the laboratory methods used
[@trotter2004; @cholongitas2007].
In addition women are disadvantaged in MELD-Na [@sealock2022].
Patients with identical disease states can have very different MELD scores and
thus receive different priority levels on the liver transplantation waiting
list.
Furthermore, for acute-on-chronic liver failure, for example, the MELD score
often underestimates the mortality risk [@hernaez2020].

@@

As shown by the MELD-Plus risk and MELD 3.0 scores, better predictive scores
often need more variables and are more complicated.
To reduce the risk of overlooking or
incorrectly calculating and interpreting the results,
clinical decision support systems may be used and could improve patient safety.
The research project on digital laboratory medicine (AMPEL) develops a
clinical decision support system based on laboratory diagnostics that should
support clinical practitioners in interpreting laboratory results and taking
the necessary medical interventions [@eckelt2020].

This study aims to find clinical and laboratory values that improve on the risk
stratification for liver transplantation of the classical
MELD, MELD-Na, MELD 3.0 and MELD-Plus scores and that might be implemented as
part of the AMPEL clinical decision support system.

@@

MELD 3.0 was calculated as described in kim2022:

$$
\begin{aligned}
MELD3 = 1.33 (if female) &+
        4.56 &* \ln(bilirubin [mg/dL]) +  \\
        0.82 &* (137 - Na [mmol/L]) - 0.24 * (137 - Na [mmol/L]) * \ln(bilirubin [mg/dL]) + \\
        9.09 &* \ln(INR) + \\
        11.14 &* \ln(creatinine [mg/dL]) + 1.85 * (3.5 - albumin [g/dL]) - 1.83 * (3.5 - albumin [g/dL]) * \ln(creatinine [mg/dL]) + 6
\end{aligned}
$$
In contrast to MELD-Na the maximum allowed creatinine was 3.0 mg/dL.

The MELD-Plus7 risk score was calculated as described in @kartoun2017:
$$
\begin{aligned}
L = 8.53499496 &+ \\
    2.59679650 &* \log_{10}(1 + creatinine [mg/dL]) + \\
    2.06503238 &* \log_{10}(1 + bilirubin [mg/dL]) + \\
    2.99724802 &* \log_{10}(1 + INR) - \\
    6.47834101 &* \log_{10}(1 + sodium [mmol/L]) - \\
    6.34990436 &* \log_{10}(1 + albumin [g/dL]) + \\
    1.92811726 &* \log_{10}(1 + wbc [th/cumm]) + \\
    0.04070442 &* age [years]
\end{aligned}
$

@@

## Data processing and machine-learning

@@

The final model was compared against the MELD, MELD-Na, MELD 3.0 and MELD-Plus7
scores using the area under the time-dependent receiver operating
characteristic curve (AUROC) based on the
nonparametric inverse probability of censoring weighting
estimator (IPCW) [@blanche2013, @R-timeROC].
Testing was applied as described in @blanche2013.
A $p < 0.05$ was considered a statistically significant difference.

@@

Most of the patients presented with cirrhosis
(`r .tbl1("Cirrhosis")`).
The remaining patients had acute liver failure
(`r .tbl1("ALF")`) or
chronic liver insufficiency
(`r .tbl1("CLI")`).
At evaluation time, the median (IQR) MELD, MELD-Na and MELD 3.0 scores were
`r .tbl1("ScoreMeldUnos")`,
`r .tbl1("ScoreMeldNaUnos")`, and
`r .tbl1("ScoreMeld3")`, respectively.
Within 90 days, `r .tbl1("M90")` patients had died.
In nearly two thirds of all patients
(`r .tbl1("Ethyltoxic")`), alcohol was the main cause
of their end-stage liver disease.
The remaining patients had viral infections -
hepatitis B (`r .tbl1("HBV")`) or C (`r .tbl1("HCV")`) -
autoimmune autoimmune (`r .tbl1("AIH")`),
primary sclerosing cirrhosis (`r .tbl1("PSC")`),
primary biliary cirrhosis (`r .tbl1("PBC")`), unknown causes
(non-alcoholic steatohepatitis (`r .tbl1("NASH")`), or
cryptogenic hepatitis (`r .tbl1("Cryptogenic")`)).
The most common complications among all patients were gastrointestinal
bleedings in `r .tbl1("GIB")` patients, followed by
hepatocellular carcinoma (HCC) in `r .tbl1("HCC")` and
spontaneous bacterial peritonitis (SBP) in `r .tbl1("SBP")` patients.
Just `r .tbl1("Dialysis")` patients required renal dialysis.

@@

We divided our cohort into
`r nlevels(labelled_meld_data$MeldCategory)` MELD-Na-score risk
categories, with `r enum(table(labelled_meld_data$MeldCategory))` patients,
respectively.
As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality),
our observed 90-day mortality was
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times higher than the predicted one [@vanderwerken2021].
Only in the group with the lowest MELD-Na scores was the mortality much lower.

```{r tbl-observed-vs-expected-mortality, results = "asis", echo = FALSE, message = FALSE}
tbl_observed_vs_expected_mortality$ObservedMortality <-
    tbl_observed_vs_expected_mortality$ObservedMortality * 100
tbl_observed_vs_expected_mortality$ExpectedMortality <-
    tbl_observed_vs_expected_mortality$ExpectedMortality * 100
knitr::kable(
    cbind(
        row.names(tbl_observed_vs_expected_mortality),
        tbl_observed_vs_expected_mortality
    ),
    row.names = FALSE,
    col.names =
        c(
            "MELD-Na category",
            "Observed deaths (n)",
            "Expected deaths (n)",
            "Standardized mortality ratio (SMR)",
            "Observed mortality (%)",
            "Expected mortality (%)"
        ),
    caption = paste0(
        "Observed vs. MELD-Na-expected 90-day mortality. ",
        "MELD-Na mortality values are taken from @vanderwerken2021 and ",
        "the mean value was calculated for each MELD-Na category.",
        "All patients censored before day 90 are ignored for ",
        "the calculation of the MELD-Na-expected deaths. ",
        "SMR, Standardized mortality ratio = observed deaths/expected deaths."
    ),
    digits = 1
)
```

@@

```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve. Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, MELD-Plus7, as described in [@blanche2013]."}
col <- viridisLite::viridis(6)
m <- list(
    AMELD = timeROC_RCV,
    MELD = timeROC_MELD,
    "MELD-Na" = timeROC_MELDNa,
    "MELD 3.0" = timeROC_MELD3,
    "MELD-Plus7" = timeROC_MELDPlus7
)
invisible(roc <- plot_surv_roc(m, timepoint = 90, col = col[-2]))
o <- order(-roc)
ci <- attr(roc, "CI")[o]
roc <- roc[o]

pAUC <- function(auc, ci) {
    sprintf("AUROC %0.3f %s", auc, ci)
}
```

@@

AMELD offers the best `r pAUC(roc["AMELD"], ci["AMELD"])` of all competitors.
However, after correction for multiple testing, the difference against
the second-best model, `r names(roc)[2]`
(`r pAUC(roc[2], ci[2])`), is not statistical significant
(`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
(see Table \@ref(tab:roctest) for complete comparison).

```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na, MELD 3.0 and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival. The error bars correspond to the 95% confidence interval of the survival estimate.'), echo = FALSE}
par(cex = 0.9)
plot(
    bootrcv,
    what = "calibration",
    main = "90-day survival",
    legend = FALSE
)

ps <- lapply(
    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "3", "Plus7"))],
    function(p) {
        ctpnts <- cutpoints(p, n = ameldcfg$m)
        f <- cut(p, ctpnts, include.lowest = TRUE)
        list(
            predicted = groupmean(p, f = f),
            observed = observed_survival(
                amelddata$y, f = f, times = ameldcfg$times
            )
        )
    }
)
names(ps) <- c("MELD", "MELD-Na", "MELD 3.0", "MELD-Plus7")

@@

In this retrospective analysis of `r n_analysed` consecutive patients
who were recruited during the evaluation process for liver transplantation at
University Hospital Leipzig, we developed a new model to predict
90-day survival probability. Our new model, AMELD,
incorporates synthesis and inflammatory markers and
could improve the prediction over previous published
MELD scores.

@@

It has been shown that women receive median lower scores in MELD-Na and
therefore are less likely to receive an urgent listing for liver transplantation
[@sealock2022].
MELD 3.0 tries to circumvent this by adding 1.33 to the score for femal sex.
To address sexual differences in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017] in our study.
This transformation also yields comparable laboratory measurements
across different clinical laboratories and methods, as well as reducing
the influence of age- and sex-related differences,
which could improve the objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample-dependent scaling prior to model development.
Both should increase the external applicability of our approach.

@@

The latter is an example where results from machine-learning can aid in
finding clinically meaningful patterns and guide further research.
Systemic inflammatory response syndrome is an
independent risk factor for poor outcomes in patients with cirrhosis
[@thabut2007].
However, neither MELD, MELD-Na nor MELD 3.0 took inflammation into account.
By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
excluded, the algorithms select CRP as one of the most important variables,
which highlights the important role of inflammation in the prognosis of
patients with end-stage liver disease.
We and other researchers have already shown that adding inflammatory markers,
namely IL-6 or CRP, to the classical MELD score increases the accuracy of the
90-day mortality prediction [@remmler2018; @cervoni2016].

@@

With a median MELD-Na score of `r .tbl1("ScoreMeldNaUnos", pattern = "{median}")`,
our cohort was healthier than those in previous studies that reported
a score of 14-18.5 [@kim2008; @kartoun2017; @vanderwerken2021].
However, the 90-day mortality rate of `r .tbl1("M90")` was comparable to
the mortality rates seen in earlier studies ranging from 6% to 30%
[@malinchoc2000; @kim2008; @kartoun2017].
Although our cohort was healthier, with a medium mortality rate, the classical
MELD score underestimates the observed mortality by
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times.
Unfortunately, our sample size was low in the high-risk patient category,
resulting in a poor but better than MELD-based prediction performance by our
AMELD for very ill patients.

@@

In conclusion, we have developed a new model to predict the 90-day survival of
patients during the evaluation process for liver transplantation.
Our model, AMELD, extends the classical MELD predictors INR,
(direct) bilirubin, and cystatin C (instead of creatinine) by using the
synthesis parameters total protein and cholinesterase and,
most importantly, the inflammatory marker IL-6.
Using these six laboratory measurements, AMELD
could improve the prediction over previous published
MELD scores.
For wider adoption, a prospective multi-center study for model
recalibration and validation is needed.

### 2022-11-30 analysis/supplement.Rmd
feat: add MELD 3.0 and adapt to reviewer's suggestions

```{r benchmark-table, echo = FALSE}
agg <- bmrk_results$aggregate(msr("surv.cindex", id = "harrell"))
agg[, `:=`
    (nr = NULL, resample_result = NULL, resampling_id = NULL, iters = NULL)
]
agg <- subset(agg, task_id == "zlog_eldd" & !grepl("^scale", learner_id))
agg[, task_id := NULL]
setorder(agg, -harrell)
agg$name <- relabel_learners(agg$learner_id)
knitr::kable(
    agg[, list(name, learner_id, harrell)],
    col.names = c("Model", "Learner ID", "Harrell C"),
    caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine-learning algorithms.")
)
```

@@

```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, and MELD-Plus7, as described in [@blanche2013]."}
tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELD3)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_RCV)

@@

>>>>

```{r benchmark-table, echo = FALSE}
agg <- bmrk_results$aggregate(msr("surv.cindex", id = "harrell"))
agg[, `:=`
    (nr = NULL, resample_result = NULL, resampling_id = NULL, iters = NULL)
]
agg <- subset(agg, task_id == "zlog_eldd" & !grepl("^scale", learner_id))
agg[, task_id := NULL]
setorder(agg, -harrell)
agg$name <- relabel_learners(agg$learner_id)
knitr::kable(
    agg[, list(name, harrell)],
    col.names = c("Model", "Harrell's concordance index"),
    caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine-learning algorithms."),
    digits = 3
)
```

# Observed versus MELD-Na-expected 90-day mortality

```{r tbl-observed-vs-expected-mortality5, results = "asis", echo = FALSE, message = FALSE}
## identical to table in article.Rmd but categories spanning just 5 points
tar_load(labelled_meld_data)

## MELD mortality values are taken from @vanderwerken2021
vanderwerken <- data.frame(
    MELD = 6:40,
    mortality = 1 -
    c(0.9700, 0.9720, 0.9715, 0.9744, 0.9802, 0.9759, 0.9747, 0.9791,
      0.9752, 0.9651, 0.9606, 0.9514, 0.9380, 0.9309, 0.9128, 0.9048,
      0.8721, 0.8509, 0.8098, 0.7869, 0.7406, 0.6787, 0.6619, 0.5791,
      0.5244, 0.4906, 0.4831, 0.4069, 0.3860, 0.3629, 0.2944, 0.2429,
      0.2469, 0.1003, 0.1562)
)

breaks <- c(-Inf, seq(10, 40, by=5), Inf)
labels <- c(
    paste0(
        "[",
        floor(min(labelled_meld_data$ScoreMeldNaUnos, na.rm = TRUE)),
        ",9]"
    ),
    paste0("[", seq(10, 35, by=5), ",", seq(15, 40, by=5), ")"),
    paste0(
        "[40,",
        ceiling(max(labelled_meld_data$ScoreMeldNaUnos, na.rm = TRUE)),
        ")"
    )
)

vanderwerken$MeldCategory <- cut(
    vanderwerken$MELD,
    breaks = breaks,
    labels = labels,
    right = FALSE
)
expected <- tapply(
    vanderwerken$mortality, vanderwerken$MeldCategory, mean
)

tbl_observed_vs_expected_mortality5 <- observed_vs_expected_mortality(
    Surv(labelled_meld_data$DaysAtRisk, labelled_meld_data$Deceased),
    f = cut(
        labelled_meld_data$ScoreMeldNaUnos,
        breaks = breaks, labels = labels,
        right = FALSE
    ),
    time = 90, expected
)

tbl_observed_vs_expected_mortality5$ObservedMortality <-
    tbl_observed_vs_expected_mortality5$ObservedMortality * 100
tbl_observed_vs_expected_mortality5$ExpectedMortality <-
    tbl_observed_vs_expected_mortality5$ExpectedMortality * 100
knitr::kable(
    cbind(
        row.names(tbl_observed_vs_expected_mortality5),
        tbl_observed_vs_expected_mortality5
    ),
    row.names = FALSE,
    col.names =
        c(
            "MELD category",
            "Observed deaths (n)",
            "Expected deaths (n)",
            "Standardized mortality ratio (SMR)",
            "Observed mortality (%)",
            "Expected mortality (%)"
        ),
    caption = paste0(
        "Observed vs. MELD-Na-expected 90-day mortality. ",
        "MELD-Na mortality values are taken from @vanderwerken2021 and ",
        "the mean value was calculated for each MELD-Na category.",
        "All patients censored before day 90 are ignored for ",
        "the calculation of the MELD-Na-expected deaths. ",
        "SMR, Standardized mortality ratio = observed deaths/expected deaths."
    ),
    digits = 1
)
```

@@

```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013]."}
tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELD3)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_RCV)

@@

```{r roc-plot-wo-cb, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013]. Identical to the figure above but without confidence bands."}
m <- lapply(m, function(x) { x$confint <- NULL;x })
plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(6)[-2])
```
